---
title: "Pertussis and the CMI-PB project"
author: "Kendrick Nguyen"
format: pdf
editor: visual
---

Pertussis is a bacterial infection that causes a severe cough. Often named "whooping cough"

Look at case numbers of pertussis in the US. We can use this website (" https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html")

```{r}
library(datapasta)
#/ echo: FALSE
cdc<- data.frame(
                                Year = c(1922L,1923L,1924L,1925L,1926L,
                                         1927L,1928L,1929L,1930L,1931L,
                                         1932L,1933L,1934L,1935L,1936L,1937L,
                                         1938L,1939L,1940L,1941L,1942L,
                                         1943L,1944L,1945L,1946L,1947L,1948L,
                                         1949L,1950L,1951L,1952L,1953L,1954L,
                                         1955L,1956L,1957L,1958L,1959L,
                                         1960L,1961L,1962L,1963L,1964L,1965L,
                                         1966L,1967L,1968L,1969L,1970L,
                                         1971L,1972L,1973L,1974L,1975L,1976L,
                                         1977L,1978L,1979L,1980L,1981L,
                                         1982L,1983L,1984L,1985L,1986L,1987L,
                                         1988L,1989L,1990L,1991L,1992L,1993L,
                                         1994L,1995L,1996L,1997L,1998L,
                                         1999L,2000L,2001L,2002L,2003L,2004L,
                                         2005L,2006L,2007L,2008L,2009L,
                                         2010L,2011L,2012L,2013L,2014L,2015L,
                                         2016L,2017L,2018L,2019L,2020L,
                                         2021L),
        No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                         202210,181411,161799,197371,166914,
                                         172559,215343,179135,265269,180518,
                                         147237,214652,227319,103188,183866,
                                         222202,191383,191890,109873,133792,
                                         109860,156517,74715,69479,120718,
                                         68687,45030,37129,60886,62786,
                                         31732,28295,32148,40005,14809,11468,
                                         17749,17135,13005,6799,7717,9718,
                                         4810,3285,4249,3036,3287,1759,
                                         2402,1738,1010,2177,2063,1623,1730,
                                         1248,1895,2463,2276,3589,4195,
                                         2823,3450,4157,4570,2719,4083,6586,
                                         4617,5137,7796,6564,7405,7298,
                                         7867,7580,9771,11647,25827,25616,
                                         15632,10454,13278,16858,27550,18719,
                                         48277,28639,32971,20762,17972,
                                         18975,15609,18617,6124,2116)
      )

```

> Q1. With the help of the R "addin" package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
library(ggplot2)
library(tidyverse)
ggplot(cdc) +
  aes(Year,No..Reported.Pertussis.Cases ) +
  geom_point() +
  geom_line() 
  
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(Year,No..Reported.Pertussis.Cases ) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, color="red") +
  geom_vline(xintercept =1996, color= "blue") +
   geom_vline(xintercept = 2019, color= "green")
  
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

cases of whooping cough started to rise meaning that people are loosing immunity to the disease. This could mean that the new aP vaccine is not as robust as the last vaccine or the bacteria is gaining resistance to the vaccines. There could be a lot of people not getting vaccinated too.

#CMI-PB project

The CMI-PB project collects and makes available data on the immune response to pertussis booster vacination

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 

head(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc...)?

```{r}
table(subject$biological_sex, subject$race)
```

```{r}
today() - mdy("08-01-1997")
time_length( today() - mdy("08-01-1997"),  "years")

```

```{r}
library(lubridate)

```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <-ymd(subject$date_of_boost)-ymd(subject$year_of_birth)

subject$age_years<- (time_length(subject$age, "years"))
```

```{r}
ggplot(subject) +
  aes(age_years, fill= infancy_vac)+
  facet_wrap(vars(infancy_vac), ncol=1)+
  geom_histogram()
```

```{r}
table(subject$dataset)
```

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
```

```{r}
head(titer)
```

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

I want to merge the speciment and subject tables together

```{r}
meta <- inner_join(specimen, subject)
dim(meta)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
head(abdata)
```

> Q11.How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q12. What are the different \$dataset values in abdata and what do you notice about the number of rows for the most "recent" dataset?

```{r}
table(abdata$dataset)
```

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

PT,PRN, FIM 2/3, FHA

```{r}
opps<- abdata %>% filter (antigen=="Fim2/3")
table(opps$dataset)
```

```{r}
table(abdata$dataset)
```

select for the 2021 dataset and isotype igg I want a time course of (planned_day_relative_to_boost) IgG MFI_Normalised) for "pT" antigen

```{r}
results<- filter(abdata, dataset=="2021_dataset",antigen=="PT", isotype=="IgG")

  ggplot(results) +
  aes(planned_day_relative_to_boost, MFI_normalised, col=infancy_vac) +
  geom_point()+
  geom_line(aes(group=subject_id), linewidth= 0.5, alpha=0.5)+
  geom_smooth(se=FALSE, span =0.4, linewidth=3)
 
```

```{r}
results1<- filter(abdata, dataset=="2022_dataset",antigen=="PT", isotype=="IgG")

 ggplot(results1) +
  aes(planned_day_relative_to_boost, MFI_normalised, col=infancy_vac) +
  geom_point()+
  geom_line(aes(group=subject_id), linewidth= 0.5, alpha=0.5)+
  geom_smooth(se=FALSE, span =0.4, linewidth=3)
```
